---
title: "Practical-1,2"
author: "Abhinav Sharma"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

The aim is to build some good linear models to analyze the ``swiss`` data where ``Fertility`` is used as the response.

## Initial Data Analysis

### Dataset Description

Standardized fertility measure and socio-economic indicators for each of 47 French-speaking provinces of Switzerland at about 1888.

### Dataset Format

A data frame with 47 observations on 6 variables, each of which is in percent, i.e., in [0, 100].

```[,1]	Fertility```	- Ig, ‘common standardized fertility measure’

```[,2]	Agriculture```	- % of males involved in agriculture as occupation

```[,3]	Examination```	- % draftees receiving highest mark on army
examination

```[,4]	Education```	- % education beyond primary school for draftees.

```[,5]	Catholic```	- % ‘catholic’ (as opposed to ‘protestant’).

```[,6]	Infant.Mortality``` - live births who live less than 1 year.

All variables but ‘Fertility’ give proportions of the population.

### Analysis

```{r}
# loading data
data(swiss)

# summary of data
summary(swiss)
```

#### 1. Distribution of response variable

```{r fig.width=6, fig.height=4}
# histogram plot
hist(swiss$Fertility,main="Fertility",xlab="Fertility")
```

```{r fig.width=6, fig.height=4}
# density plot
plot(density(swiss$Fertility),main="Fertility")
rug(swiss$Fertility)
```

- Plot is little right skewed
- Fertility rates are mostly between 60-90%

#### 2. Analyzing predictor variables

```{r fig.width=8, fig.height=6}
# boxplot analysis
boxplot(swiss)
```

- ``Catholic`` variable covers wide range of values
- ``Infant.Mortality`` variable is very condensed
- ``Education`` seems to have some outliers

```{r}
# correlation analysis
cor(swiss)
```

- All correlations with ```Fertility``` are less than 0.8, indicating no signs of strong multicollinearity.
- Correlations are betweeen 0.3-0.8, indicating mild multicollinearity.

```{r}
# correlation analysis plot
pairs(swiss)
```

- Plot shows linear relationship between ```Agriculture``` and ```Examination```.
- Also, between ```Examination``` and ```Education```.
- Interpretation of coefficients will be affected.

## Fitting a linear Model

```{r}
# basic linear model with all variables
model <- lm(Fertility ~ Agriculture + Examination + Education + 
              Catholic + Infant.Mortality, swiss)
summary(model)
```

## Transformations

We can use box-cox transformation method to understand if response variable transformation is needed or not.

```{r}
# box-cox transformation for response variable
library(MASS)
box <- boxcox(
Fertility ~ Agriculture + Examination +Education+ Catholic +
Infant.Mortality, data = swiss,
lambda = seq(from = -3, to = 3, length = 50)
)
```

- $\lambda$=1 lies in 95% confidence interval, so we can say that no transformation is needed for response variable.

Transformations of predictors can also be achieved - ?? Maybe polynomials ??

## Variable Selection

```{r}
# stepwise selection using AIC
model2 <- step(model,trace=FALSE)
model2
```

- The final AIC of model achieved = 189.86
- Final Model: ```Fertility ~ Agriculture + Education + Catholic + Infant.Mortality```

```{r}
# backward elimination using F-test
drop1(model,test='F')
```

- Backward elimination using F-test also suggest removal of ``Examination`` term as p-value is extremely large (0.315462 > 0.05).
- Final Model is same as selected using stepwise.

## Diagnostics

```{r fig.height=7}
# Basic diagnostic plots
par(mfrow=c(2,2))
plot(model2)
```

- From residual plot, there is no pattern observed so constant variance assumption holds.

- From QQ plot, we don't see heavy tails so normality assumption also holds.

- Some extreme values are observed in cook's distance plot, indicating outliers and influential points in the data.


```{r}
# top influential points
swiss[cooks.distance(model2) > 0.1,]
```

```{r}
# possible outliers
library(faraway)
halfnorm(influence(model2)$hat)
swiss[influence(model2)$hat>0.3,]
```

- We can handle this by either removing outliers points from the dataset and then proceed with least squares.

- We can also perform ``robust regression`` that downweights the effects of larger errors.


```{r}
# Robust regression
library(MASS)
rlmodel <- rlm(Fertility ~ Agriculture + Examination + Education + 
              Catholic + Infant.Mortality, swiss)
summary(rlmodel)
```

## Predictions

Suppose we assume following data for a particular region:

```[,2]	Agriculture```	- 70 % of males involved in agriculture as occupation

```[,3]	Examination```	- 20 % draftees receiving highest mark on army
examination

```[,4]	Education```	- 30 % education beyond primary school for draftees.

```[,5]	Catholic```	- 10 % ‘catholic’ (as opposed to ‘protestant’).

```[,6]	Infant.Mortality``` - 28 live births who live less than 1 year

```{r}
# input data
data <- data.frame(Agriculture=70,Examination=20,Education=30,
                   Catholic=10,Infant.Mortality=28)
# predict fertility rate for a region using full model
predict(model, newdata=data)
```

```{r}
# predict fertility rate using stepwise model
predict(model2, newdata=data)
```

```{r}
# predict fertility rate using robust regression model
predict(rlmodel, newdata=data)
```

## Interpretation

Let's use ``summary`` statistics of different variables to get different values of predictions and then interpret results for ``fertility``.

```{r}
# input data
data <- data.frame(
  Agriculture=c(1.20,35.90,50.66,67.95,89.70),
  Examination=c(3,12,16.49,22,37),
  Education=c(1,6,10.98,12,53),
  Catholic=c(2.150,5.195,41.144,93.125,100),
  Infant.Mortality=c(10.80,18.15,19.94,21.70,26.60)
)
# using full model to get complete interpretation
data$pp <- predict(model,newdata=data)
matplot(data,type='o',ylab='Fertility',xlab='Variables')
```

Here:

- ```1-Agriculture```
- ```2-Examination```
- ```3-Education```
- ```4-Catholic```
- ```5-Infant.Mortality```
- ```6-Fertility```

Interpretation:

- Fertility goes down as other predictor variables % increases
- Fertility is highly affected by ```Agriculture``` and ```Catholic``` variables.
